---
title: "Cleaning data"
format: html
---

```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(patchwork) # combine plots
library(labelled)  # For adding and extracting labels
```

# Quality checks and removing NAs

Read data.
```{r, message=FALSE}
# uses the csv file with numeric values as exported from Qualtrics
d <- read_csv("data/qualtrics.csv")
```

```{r, include=FALSE}
# inspect data
names(d)
head(d)
```

Delete empty observations in the first two rows. 

```{r}
# delete first two rows
d <- d %>% slice(3:n())
```

## Attention checks

First attention check

```{r, echo=FALSE}
# attention check number 1:

# to see different answers given (i.e.levels), transform into factor
d$attcheck_number <- as.factor(d$ATTCHECK_NUMBER)
# check levels to see different answer types
levels(d$attcheck_number) 
knitr::kable(table(d$ATTCHECK_NUMBER, useNA = "always"))
```

Second attention check

```{r}
# attention check number 2:

# to see different answers given (i.e.levels), transform into factor
d$attcheck_res <- as.factor(d$ATTCHECK_RES)
# check levels to see different answer types
levels(d$attcheck_res) 
knitr::kable(table(d$ATTCHECK_RES, useNA = "always"))
```


```{r, echo=FALSE}
# make plot data
plot_data_general <- d %>% 
  mutate(
    # create attention check 'failed' / 'passed` 
    attcheck_number = tolower(attcheck_number),
    attcheck_number = ifelse(
    # make sure to count correct answers written in text as 'passed', too.
    attcheck_number == "213" | str_detect(attcheck_number, "treize"), 
    "passed", "failed"), 
    attcheck_res  = ifelse(attcheck_res == 1, "passed", 
                           "failed"), 
    attention_checks = ifelse(attcheck_res == "failed" | attcheck_number == "failed", 
                              "failed", "passed"), 
    # code NAs as failed
    attention_checks = ifelse(is.na(attention_checks), "failed", attention_checks)
    )

plot_data_detail <- plot_data_general %>% 
  pivot_longer(starts_with("attcheck", ignore.case = FALSE), 
               names_to = "attcheck_type", 
               values_to = "attcheck_outcome") %>% 
  # make sure NAs are coded as 'failed'
  mutate(attcheck_outcome = ifelse(is.na(attcheck_outcome), "failed", 
                                   attcheck_outcome)
         ) %>% 
  group_by(attcheck_type, attcheck_outcome) %>% 
  summarise(n = n_distinct(ResponseId))

# make graphs to vizalize
attention_check_general <- ggplot(plot_data_general, aes(x = attention_checks, fill = attention_checks)) + 
  geom_bar(position = "dodge") +
  geom_text(aes(label = ..count..), stat = "count", position = position_dodge(width = 0.9), vjust = -0.1)  + 
  labs(title = "Overall", x = NULL)

attention_check_details <- ggplot(plot_data_detail, aes(x = attcheck_type, y= n,  fill = attcheck_outcome)) + 
  geom_col(position = "dodge") + 
  coord_flip() +
  geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = 0.1) +
  labs(title = "Per attention check", x = NULL)

ggpubr::ggarrange(attention_check_general, attention_check_details, common.legend = TRUE, legend = "bottom")
```

```{r}
# delete failed attention checks
d <- d %>%
  mutate(
    attcheck_number = tolower(attcheck_number),
    attcheck_number =
      ifelse(attcheck_number == "213" | str_detect(attcheck_number, "treize"),
             attcheck_number, NA),
    attcheck_res  = ifelse(attcheck_res == 1,
                           attcheck_res, NA)
  )

d <- d %>% 
  drop_na(attcheck_number, attcheck_res)
```

There also appears to be one preview run included.
```{r}
table(d$DistributionChannel, useNA = "always")
```

We remove this run

```{r}
d <- d %>% 
  mutate(DistributionChannel = ifelse(DistributionChannel == "preview", NA, DistributionChannel)) %>% 
  drop_na(DistributionChannel)
```

### Response time.

```{r, include=FALSE, echo=FALSE}
# calculate response time
time <- d %>% 
  select(ResponseId, StartDate, EndDate) %>% 
  mutate(time = difftime(EndDate, StartDate, units="mins"))

# get measures of central tendency
mean(time$time)
median(time$time) 
```

```{r, echo=FALSE}
# inspect graphically
ggplot(time, aes(x = time)) +
  geom_histogram() +
  geom_vline(aes(xintercept = mean(time, na.rm=TRUE)), colour = "red") +
  labs(title = "Time among respondents who passed \n all attention checks ", 
       x = "time (in mins)") +
  annotate("text", x = 2000, y = 300, label = paste("n =", nrow(time)))
```

```{r, echo=FALSE}
# filter time
time_reduced <- time %>% filter(time <= 60)

# check without outlier
ggplot(time_reduced, aes(x = time)) +
  geom_histogram() +
  geom_vline(aes(xintercept = mean(time, na.rm=TRUE)), colour = "red") + 
  scale_x_continuous(breaks = seq(from = 0, to = 60, by = 5)) +
  labs(title = "Time among respondents who passed \n all attention checks, removing those who took more than 1h", 
       x = "time (in mins)") +
  annotate("text", x = 40, y = 40, label = paste("n =", nrow(time_reduced)))
```

### Response date

```{r, echo=FALSE}
date <- ggplot(d, aes(x = as.factor(as.Date(StartDate)))) + 
  geom_bar(position = "dodge") +
  geom_text(aes(label = ..count..), stat = "count", position = position_dodge(), vjust = -0.1)  + 
  labs(title = "Start date of survey, 
       valid (i.e. passed attention checks) responses only", x = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

date
```

### Age

```{r, echo=FALSE}
# inspect graphically
ggplot(d, aes(x = as.numeric(DEM_AGE))) +
  geom_histogram() +
  scale_x_continuous(breaks = seq(from = 0, to = 100, by = 5)) +
  geom_vline(aes(xintercept = mean(as.numeric(DEM_AGE), na.rm=TRUE)), colour = "red") +
  labs(title = "Age of respondents who passed \n all attention checks ", 
       x = "age") +
  annotate("text", x = 75, y = 100, label = paste("n =", nrow(d)))
```

# Specific checks for additional questions

We recruited a total sample of ~2000 participants together as team France. Each of the 4 binomes added a set of questions. However, to not make the survey too long, we split our additional questions: Half of participants would be randomly assigned to one set of questions, the other half to the other.

For our variables, check whether variable names match questions. 

```{r, echo=FALSE, include=FALSE, message=FALSE}
# check if questions correspond to variable name for a selection of variables
knitr::kable(
  # we have to read in the original data again since we deleted the first
  # row with the question framing already
  read_csv("data/qualtrics.csv") %>% 
               select(starts_with("disciplines"), subjective_status) %>% 
               slice(1) %>% 
               pivot_longer(everything(), 
                            names_to = "variable", 
                            values_to = "question"), 
             "html") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

### Scientific disciplines

Check how many observations we have.

```{r, include=FALSE, echo=FALSE}
# Disciplines NA check
disciplines_NA <- d %>% 
  pivot_longer(cols = starts_with("disciplines"), names_to = "variable_discipline", 
               values_to = "score", names_prefix = "disciplines_") %>% 
  group_by(ResponseId) %>% 
  summarise(sum_NA = sum(is.na(score))) 

table(disciplines_NA$sum_NA)
```

```{r, echo=FALSE}
disciplines_NA_plot <- ggplot(disciplines_NA, aes(x = as.factor(sum_NA))) + 
  geom_bar(position = "dodge") +
  geom_text(aes(label = ..count..), stat = "count", position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Number of NAs across the 18 discipline questions", 
       x = "Total sum of NA's across all 18 questions")

disciplines_NA_plot
```

We have 1008 valid answers and three people who have 1 NA value, and 1 person who has 13 NA values. 

### Subjective status

Check how many observations we have.

```{r, include=FALSE, echo=FALSE}
# Status NA check
status_NA <- d %>% group_by(ResponseId) %>% 
  summarize(sum_NA = sum(is.na(subjective_status))) %>% 
    select(sum_NA) %>% 
  mutate(sum_NA = ifelse(sum_NA == 1, "NA", "No NA"))

table(status_NA$sum_NA)
```

```{r, echo=FALSE}
status_NA_plot <- ggplot(status_NA, aes(x = sum_NA,  fill = sum_NA)) + 
  geom_bar(position = "dodge") +
  geom_text(aes(label = ..count..), stat = "count", position = position_dodge(width = 0.9), vjust = - 0.1) +
  labs(title = "Number of NAs vs. No NAs for status question", x = NULL) + 
  guides(fill = FALSE)

status_NA_plot
```

For the status question, we have 1015 valid answers. 

# Data cleaning

```{r}
# give it a nicer 'id' column
d <- d %>% mutate(id = 1:nrow(d))
```

### Select valid participants for discipline questions

Among the total sample, only half were randomly selected to answer the questions about the different scientific disciplines. We therefor limit the sample to those participants who saw these question. 

As shown above, we want to exclude participants who have only NA responses. 

```{r}
# select participants to exclude
participants_to_exclude <- d %>% 
  pivot_longer(cols = starts_with("disciplines"), names_to = "variable_discipline", 
               values_to = "score", names_prefix = "disciplines_") %>% 
  group_by(ResponseId) %>% 
  summarise(sum_NA = sum(is.na(score))) |> 
  filter(sum_NA == 18) |> 
  pull(ResponseId)

# check
length(participants_to_exclude)

# subset the data
d <- d |> filter(!ResponseId %in% participants_to_exclude)
```

### Benevolence of scientists

The main survey has three items intended to capture the factor `benevolence`:

-   `benevolence_concerned`(original survey coding: `TRUST_SCI_concerned`): "How concerned or not concerned are most scientists about people's wellbeing?" [1 = Not concerned, 2 = Somewhat not concerned, 3 = Neither concerned nor not concerned, 4 = Somewhat concerned, 5 = Very concerned]

-   `benevolence_improve`(original survey coding: `TRUST_SCI_improve`): "How eager or uneager are most scientists to improve others' lives?" [1 = Very uneager, 2 = Somewhat uneager, 3 = Neither eager nor uneager, 4 = Somewhat eager, 5 = Very eager]

-   `benevolence_otherinterest`(original survey coding: `TRUST_SCI_otherint`): "How considerate or inconsiderate are most scientists of others' interests?" [1 = Very inconsiderate, 2 = Somewhat inconsiderate, 3 = Neither considerate nor inconsiderate, 4 = Somewhat considerate, 5 = Very considerate]

```{r}
# do the re-naming and transform to numeric
d <- d %>%
  # syntax for rename() is new_name = old_name
  rename(benevolence_concerned = TRUST_SCI_concerned, 
         benevolence_improve = TRUST_SCI_improve,
         benevolence_otherinterest = TRUST_SCI_otherint
         ) %>% 
  # For now, all variables containing numbers are coded as 'character'.
  # We need them as numeric.
  mutate(across(starts_with("benevolence"), as.numeric))
```


We will treat them as three different measures of the same construct, `benevolence` in our analysis. Practically, this means that for our analysis, we will simply calculate their mean as our measure of `benevolence`.

```{r}
d <- d %>%
  # calculate composite measure
  mutate(benevolence = rowMeans(across(starts_with('benevolence')), na.rm = TRUE)
         )
```

We assign labels.

```{r}
# Assign variable labels (questions)
d <- d %>%
  set_variable_labels(
    benevolence_concerned = "How concerned or not concerned are most scientists about people's wellbeing?",
    benevolence_improve = "How eager or uneager are most scientists to improve others' lives?",
    benevolence_otherinterest = "How considerate or inconsiderate are most scientists of others' interests?"
  )

# Assign value labels (answer options)
d <- d %>%
  set_value_labels(
    benevolence_concerned = c("Not concerned" = 1, "Somewhat not concerned" = 2, 
                              "Neither concerned nor not concerned" = 3, "Somewhat concerned" = 4, "Very concerned" = 5),
    
    benevolence_improve = c("Very uneager" = 1, "Somewhat uneager" = 2, 
                            "Neither eager nor uneager" = 3, "Somewhat eager" = 4, "Very eager" = 5),
    
    benevolence_otherinterest = c("Very inconsiderate" = 1, "Somewhat inconsiderate" = 2, 
                                  "Neither considerate nor inconsiderate" = 3, "Somewhat considerate" = 4, "Very considerate" = 5)
  )
```


### Competence of scientists

The main survey has three items intended to capture the factor `competence`:

-   `competence_expert`(original survey coding:`TRUST_SCI_expert`): "How expert or inexpert are most scientists?" [1 = Very inexpert, 2 = Somewhat inexpert, 3 = Neither expert nor inexpert, 4 = Somewhat expert, 5 = Very expert]

-   `competence_intelligent`(original survey coding:`TRUST_SCI_intellig`): "How intelligent or unintelligent are most scientists?" [1 = Very unintelligent, 2 = Somewhat unintelligent, 3 = Neither intelligent nor unintelligent, 4 = Somewhat intelligent, 5 = Very intelligent]

-   `competence_qualified`(original survey coding:`TRUST_SCI_qualified`): "How qualified or unqualified are most scientists when it comes to conducting high-quality research?" [ 1 = Very unqualified, 2 = Somewhat unqualified, 3 = Neither qualified nor unqualified, 4 = Somewhat qualified, 5 = Very qualified]

```{r}
# do the re-naming and transform to numeric
d <- d %>%
  # syntax for rename() is new_name = old_name
  rename(competence_expert = TRUST_SCI_expert, 
         competence_intelligent = TRUST_SCI_intellig,
         competence_qualified = TRUST_SCI_qualified
         ) %>% 
  # For now, all variables containing numbers are coded as 'character'.
  # We need them as numeric.
  mutate(across(starts_with("competence"), as.numeric))
```

We will treat them as three different measures of the same construct, `competence` in our analysis. Practically, this means that for our analysis, we will simply calculate their mean as our measure of `competence`.

```{r}
d <- d %>%
  # calculate composite measure
  mutate(competence = rowMeans(across(starts_with('competence')), na.rm = TRUE)
         )
```

We assign labels. 

```{r}
# Assign variable labels (questions)
d <- d %>%
  set_variable_labels(
    competence_expert = "How expert or inexpert are most scientists?",
    competence_intelligent = "How intelligent or unintelligent are most scientists?",
    competence_qualified = "How qualified or unqualified are most scientists when it comes to conducting high-quality research?"
  )

# Assign value labels (answer options)
d <- d %>%
  set_value_labels(
    competence_expert = c("Very inexpert" = 1, "Somewhat inexpert" = 2, 
                          "Neither expert nor inexpert" = 3, "Somewhat expert" = 4, "Very expert" = 5),
    
    competence_intelligent = c("Very unintelligent" = 1, "Somewhat unintelligent" = 2, 
                               "Neither intelligent nor unintelligent" = 3, "Somewhat intelligent" = 4, "Very intelligent" = 5),
    
    competence_qualified = c("Very unqualified" = 1, "Somewhat unqualified" = 2, 
                             "Neither qualified nor unqualified" = 3, "Somewhat qualified" = 4, "Very qualified" = 5)
  )

```


### Integrity of scientists

The main survey has three items intended to capture the factor `integrity`:

-   `integrity_honest`(original survey coding:`TRUST_SCI_honest`): "How honest or dishonest are most scientists?" [1 = Very dishonest, 2 = Somewhat dishonest, 3 = Neither honest nor dishonest, 4 = Somewhat honest, 5 = Very honest]

-   `integrity_ethical`(original survey coding:`TRUST_SCI_ethical`): "How ethical or unethical are most scientists?" [1 = Very unethical, 2 = Somewhat unethical, 3 = Neither ethical nor unethical, 4 = Somewhat ethical, 5 = Very ethical]

-   `integrity_sincere`(original survey coding:`TRUST_SCI_sincere`): "How sincere or insincere are most scientists?" [ 1 = Very insincere, 2 = Somewhat insincere, 3 = Neither sincere nor insincere, 4 = Somewhat sincere, 5 = Very sincere]

```{r}
# do the re-naming and transform to numeric
d <- d %>%
  # syntax for rename() is new_name = old_name
  rename(integrity_honest = TRUST_SCI_honest, 
         integrity_ethical = TRUST_SCI_ethical,
         integrity_sincere = TRUST_SCI_sincere
         ) %>% 
  # For now, all variables containing numbers are coded as 'character'.
  # We need them as numeric.
  mutate(across(starts_with("integrity"), as.numeric))
```

We will treat them as three different measures of the same construct, `integrity` in our analysis. Practically, this means that for our analysis, we will simply calculate their mean as our measure of `integrity`.

```{r}
d <- d %>%
  # calculate composite measure
  mutate(integrity = rowMeans(across(starts_with('integrity')), na.rm = TRUE)
         )
```

We assign labels. 

```{r}
# Assign variable labels (questions)
d <- d %>%
  set_variable_labels(
    integrity_honest = "How honest or dishonest are most scientists?",
    integrity_ethical = "How ethical or unethical are most scientists?",
    integrity_sincere = "How sincere or insincere are most scientists?"
  )

# Assign value labels (answer options)
d <- d %>%
  set_value_labels(
    integrity_honest = c("Very dishonest" = 1, "Somewhat dishonest" = 2, 
                         "Neither honest nor dishonest" = 3, "Somewhat honest" = 4, "Very honest" = 5),
    
    integrity_ethical = c("Very unethical" = 1, "Somewhat unethical" = 2, 
                          "Neither ethical nor unethical" = 3, "Somewhat ethical" = 4, "Very ethical" = 5),
    
    integrity_sincere = c("Very insincere" = 1, "Somewhat insincere" = 2, 
                          "Neither sincere nor insincere" = 3, "Somewhat sincere" = 4, "Very sincere" = 5)
  )
```


### Openness of scientists

The main survey has three items intended to capture the factor `openness`:

-   `openness_open`(original survey coding:`TRUST_SCI_open`): "How open are most scientists to feedback??" [1 = Not open, 2 = Somewhat open, 3 = Neither open or not open, 4 = Somewhat open, 5 = Very open]

-   `openness_transparent`(original survey coding:`TRUST_SCI_trans`): "How willing or unwilling are most scientists to be transparent?" [1 = Very unwilling, 2 = Somewhat unwilling, 3 = Neither willing nor unwilling, 4 = Somewhat willing, 5 = Very willing]

-   `openness_otherviews`(original survey coding:`TRUST_SCI_otherviews`): "How much or little attention do scientists pay to others' views?" [ 1 = Very little attention, 2 = Somewhat little attention, 3 = Neither much nor little attention, 4 = Somewhat much attention, 5 = Very much attention]

```{r}
# do the re-naming and transform to numeric
d <- d %>%
  # syntax for rename() is new_name = old_name
  rename(openness_open = TRUST_SCI_open, 
         openness_transparent = TRUST_SCI_trans,
         openness_otherviews = TRUST_SCI_otherviews
         ) %>% 
  # For now, all variables containing numbers are coded as 'character'.
  # We need them as numeric.
  mutate(across(starts_with("openness"), as.numeric))
```

We will treat them as three different measures of the same construct, `openness` in our analysis. Practically, this means that for our analysis, we will simply calculate their mean as our measure of `openness`.

```{r}
d <- d %>%
  # calculate composite measure
  mutate(openness = rowMeans(across(starts_with('openness')), na.rm = TRUE)
         )
```

Next, we add labels to our data.

```{r}

# Assign variable labels (questions)
d <- d %>%
  set_variable_labels(
    openness_open = "How open are most scientists to feedback?",
    openness_transparent = "How willing or unwilling are most scientists to be transparent?",
    openness_otherviews = "How much or little attention do scientists pay to others' views?"
  )

# Assign value labels (answer options)
d <- d %>%
  set_value_labels(
    openness_open = c("Not open" = 1, "Somewhat open" = 2, 
                      "Neither open nor not open" = 3, "Somewhat open" = 4, "Very open" = 5),
    
    openness_transparent = c("Very unwilling" = 1, "Somewhat unwilling" = 2, 
                             "Neither willing nor unwilling" = 3, "Somewhat willing" = 4, "Very willing" = 5),
    
    openness_otherviews = c("Very little attention" = 1, "Somewhat little attention" = 2, 
                            "Neither much nor little attention" = 3, "Somewhat much attention" = 4, "Very much attention" = 5)
  )
```

### Influence (normative) of scientists on policy

The main survey has six items intended to capture perceptions "normative perceptions" on the role of scientists.

-   `influence_integrate` (original survey coding: `NORMPERC_integrate`): "Scientists should work closely with politicians to integrate scientific results into policy-making." [1 = strongly disagree, 5 = strongly agree]

-   `influence_advocate` (original survey coding: `NORMPERC_advocate`): "Scientists should actively advocate for specific policies." [1 = strongly disagree, 5 = strongly agree]

-   `influence_communicate` (original survey coding: `NORMPERC_communicate`): "Scientists should communicate their findings to politicians." [1 = strongly disagree, 5 = strongly agree]

-   `influence_involved` (original survey coding: `NORMPERC_involved`): "Scientists should be more involved in the policy-making process." [1 = strongly disagree, 5 = strongly agree]

-   `outreach` (original survey coding: `NORMPERC_outreach`): "Scientists should communicate about science with the general public." [1 = strongly disagree, 5 = strongly agree]

-   `influence_independent` (original survey coding: `NORMPERC_independent`): "Scientists should remain independent from the policy-making process." [1 = strongly disagree, 5 = strongly agree]

We will use five of these items to capture perceptions on the influence scientists should have on policy making (all but `outreach`). 

```{r}
# do the re-naming and transform to numeric
d <- d %>%
  # syntax for rename() is new_name = old_name
  rename(influence_integrate = NORMPERC_integrate, 
         influence_advocate = NORMPERC_advocate,
         influence_communicate = NORMPERC_communicate,
         influence_involved = NORMPERC_involved,
         influence_independent = NORMPERC_independent
         ) %>% 
  # For now, all variables containing numbers are coded as 'character'.
  # We need them as numeric.
  mutate(across(starts_with("influence"), as.numeric))
```

Precisely, we will build a composite measure `influence_policy` taking their mean.

```{r}
d <- d %>%
  # calculate composite measure
  mutate(influence_policy = rowMeans(across(starts_with('influence')), na.rm = TRUE)
         )
```

### Influence (normative) of scientists on private lifestyle

The main survey has several items intended to capture "willingness to be vulnerable" towards scientists. One items captures influence scientists should have on private lifestyle.

-   `influence_lifestyle` (original survey coding: `WILLVUL_ppl`): "How much or little should people rely on scientists' guidance when making lifestyle choices related to science?" [1 = not at all, 5 = very strongly]

```{r}
d <- d %>% 
  # do the re-naming and transform to numeric
  mutate(influence_lifestyle = as.numeric(WILLVUL_ppl)) %>% 
  # delete old variable
  select(-c(WILLVUL_ppl))
```

### Trust, consensus and precision by discipline

We added a set of questions to the main survey where we asked participants to give distinct answers for the following set of different scientific disciplines (including one 'general' item):

-   Science in general
-   Biology
-   Sociology
-   Economics
-   Physics
-   Climate science

The questions are: 

-   `disciplines_trust` : "How much do you trust the researchers working in these disciplines?" [1 = Not trust at all, 5 = Trust very much]

-   `disciplines_consensus` : "How much do you think researchers in these disciplines agree on fundamental findings?" [1 = They do not agree at all, 5 = They agree very much]

-   `disciplines_precision` : "How precise do you think the results obtained by researchers in these disciplines are?" [1 = Not precise at all, 5 = Extremely precise]

```{r}
# no need to rename, but transform to numeric
d <- d  %>% 
  mutate(across(starts_with("disciplines"), as.numeric))
```

### Subjective social status

We added a question regarding subjective social status to the main survey. The question is from the US General Social Survey, but we reversed the coding for the scale (such that 10 = top) and also let the scale run from bottom (left) to top (right) since that was consistent with all other scales in the survey.

-   `subjective_status` (coding in the GSS: `rank`) : "In our society there are groups which tend to be towards the top and groups which tend to be towards the bottom. Below is a scale that runs *from bottom to top*. Where would you put yourself on this scale?" [10 = Top, 1 = Bottom]

```{r}
# no need to rename, but transform to numeric
d <- d  %>% 
  mutate(across(subjective_status, as.numeric))
```

### Trust in scientific method

The survey includes one question on trust in the scientific method.

-   `trust_method`(original survey coding:`TRUST_METHOD`): "To what extent do you agree or disagree with the following statement? 'Scientific research methods are the best way to find out if something is true or false.'" [1 = Strongly disagree, 5 = Strongly agree]

```{r}
# do the re-naming and transform to numeric
d <- d %>%
  rename_with(tolower, matches("TRUST_METHOD")) %>% 
  # For now, all variables containing numbers are coded as 'character'.
  # We need them as numeric.
  mutate(across(trust_method, as.numeric))
```

### Trust in scientists 

Based on the measures for `competence`, `benevolence`, `integrity` and `openness`, we build a composite variable `trust_composite`. Practically, this means that for our analysis, we will simply calculate the mean of all 12 items.

```{r}
d <- d %>%
  # calculate composite measure
  mutate(trust_composite = rowMeans(
    across(starts_with(c("benevolence", "competence", "integrity", "openness")) & contains("_")), 
    na.rm = TRUE)
  )
```

To be able to run our pre-registered analysis, we also add a reduced composite measure that is only based on `competence` and `benevolence`. 

```{r}
d <- d %>%
  # calculate composite measure
  mutate(trust_composite_reduced = rowMeans(
    across(starts_with(c("benevolence", "competence")) & contains("_")), 
    na.rm = TRUE)
  )
```

### Education

-   `education`(original survey coding:`DEM_EDU`): "What is your highest completed level of education?" [1 = Did not attend school, 2 = Primary education, 3 = Secondary education (e.g., high school), 4 = Higher education (e.g., university degree or higher education diploma)]

```{r}
# do the re-naming and transform to numeric
d <- d %>%
  # make a numeric version
  mutate(education_numeric = as.numeric(DEM_EDU)) %>% 
  # make a factor variable version 
  mutate(
    education = factor(education_numeric, 
                       levels = 1:4, 
                       labels = c("Did not attend school", 
                                  "Primary education", 
                                  "Secondary education (e.g., high school)", 
                                  "Higher education (e.g., university degree or higher education diploma)"))
  )
```

### Gender

- `DEM_GENDER`: 0 = "Woman", 1 = "Man", 2 = "Prefer to self-describe", 99 = "Prefer not to say"

```{r}
# code a new gender variable
d <- d %>%
  mutate(
    gender = factor(DEM_GENDER, 
                    levels = c(0, 1, 2, 99), 
                    labels = c("Woman", "Man", "Prefer to self-describe", "Prefer not to say"))
  )
```

### Age

```{r}
d <- d %>%
  mutate(age = as.numeric(DEM_AGE))
```

### Religion

- `DEM_RELIGIOUS` "Please indicate to what extent you consider yourself religious." [Not religious at all  (1) - Very strongly religious  (5)]

```{r}
d <- d %>%
  mutate(religious = factor(
    DEM_RELIGIOUS,
    levels = 1:5,  # Ensure correct ordering
    labels = c("Not religious at all (1)", "(2)", 
               "(3)", "(4)", 
               "Very strongly religious (5)")
  ))
```

### Urban

- `DEM_RESIDENCE` "Which of the following best describes the area you live in?" [Rural (1), Urban (2)]

```{r}
d <- d %>%
  mutate(residence = factor(
    DEM_RESIDENCE,
    levels = 1:2,  # Ensure correct ordering
    labels = c("Rural", 
               "Urban")
  ))

# check
# d |> 
#   select(residence, DEM_RESIDENCE)
```


### Income

-   `income`(original survey coding:`DEM_INCOME`): "Please indicate your household's yearly net income (in local currency)."

This is a sensitive question in France and it's worth checking how well people responded to this question. 

```{r}
# Before transforming, check all unique values. 
d %>% select(DEM_INCOME) %>% 
  group_by(DEM_INCOME) %>% 
  count() 
```

Also, some people used dots, but not as correct delimiters. 

```{r}
d %>% 
  select(DEM_INCOME) %>% 
  group_by(DEM_INCOME) %>% 
  count() %>%
  filter(str_detect(DEM_INCOME, "\\.")) %>% 
  mutate(DEM_INCOME_numeric = as.numeric(str_remove(DEM_INCOME, "\\.")))
```

Only in one case, the "20000.00" the delimiter appears to have been used correctly. 

First, we transform all values to numeric.

```{r}
# do the re-naming and transform to numeric
d <- d %>%
  # we keep both variables to be able to cross-check
  mutate(income = case_when(
    # for the one case, don't remove the dot
    DEM_INCOME == "20000.00" ~ as.numeric(DEM_INCOME), 
    # for all other cases, remove it
    TRUE ~ as.numeric(str_remove(DEM_INCOME, "\\.")
    )
  )
  ) 
```

Check values

```{r}
ggplot(d, aes(income)) +
  geom_histogram(bins = 30)
```

Clearly, at least some values are unrealistically low, others unrealistically high. 

Let's inspect the individuals with an income of more than `100'000` in more detail. 

```{r, echo=FALSE}
d %>% filter(income > 100000) %>% 
  select(income, DEM_INCOME) %>% 
  arrange(desc(income))
```

The two largest values are obviously invalid responses. We code them as `NA`. 

```{r}
# do the re-naming and transform to numeric
d <- d %>%
  # code two obvious outliers as `NA`
  mutate(income = ifelse(income > 1000000, NA, income))
```

Then we can distinguish some income groups.  

```{r, echo=FALSE}
# check how many we remove by accepting only <= one million as valid income
d %>% mutate(income_categories = case_when(income >100000 ~ "more than 100'000",
                                              income == 0 ~ "zero",
                                              income <10000 ~ "less than 10'000 (but not zero)", 
                                              is.na(income) ~ NA_character_, 
                                              TRUE ~ "somewhere in between"
                                              )
                  ) %>% group_by(income_categories) %>% 
  count()
```

```{r, echo=FALSE}
# plot income distributions
plot_data <- d %>% filter(income < 10000)
low_income <- ggplot(plot_data, aes(x = income)) +
  geom_density() +
  labs(title = "Distribution for \n income < 10'000") +
  annotate("text", x = 7500, y = 0.00015, label = paste("n =", nrow(plot_data)))


plot_data <- d %>% filter(income > 100000)
high_income <- ggplot(plot_data, aes(x = income)) +
  geom_density() +
  labs(title = "Distribution for \n income > 100'000") +
  annotate("text", x = 500000, y = 0.000003, label = paste("n =", nrow(plot_data)))


plot_data <- d %>% filter(!is.na(income))
income <- ggplot(d, aes(x = income)) +
  geom_density() +
  labs(title = "Overall distribution for income") +
  annotate("text", x = 500000, y = 0.00001, label = paste("n =", nrow(plot_data)))


income / (low_income + high_income)
```

We did not specify this in our pre-registration, but due to the shape of the income distribution, we will also build a log-transformed income variable, `income_log`.

```{r}
# add a log-transformed version of `income`
d <- d %>% 
  mutate(income_log = log(income))
```

```{r, echo=FALSE}
# plot logged income
plot_data <- d %>% filter(!is.na(income_log))
income <- ggplot(d, aes(x = income_log)) +
  geom_density() +
  labs(title = "Overall distribution for log income") +
  annotate("text", x = 5, y = 0.3, label = paste("n =", nrow(plot_data)))
income
```

# Select target variables

# Long format version

In order to compare different disciplines, we need to bring our data into a long format version. 

```{r}
# make long-format data 
data_long <- d %>% 
  pivot_longer(cols = starts_with("disciplines"), names_to = "variable_discipline", 
               values_to = "score", names_prefix = "disciplines_") %>% 
  # separate variable and discipline into two distinct columns
  separate_wider_delim(variable_discipline, delim = "_", 
                       names = c("variable", "discipline")) %>% 
  # pivot wider variable and score 
  # (such that three distinct columns for trust, consensus and precision)
  pivot_wider(names_from = variable, values_from = score)
```

This makes that instead of having this... 

```{r, echo=FALSE}
# wide format
knitr::kable(d %>%  
               filter(id == 4) %>% 
               select(id, starts_with("disciplines_consensus", ignore.case = FALSE)),
             caption = "Wide format data")
```

We have this...

```{r, echo=FALSE}
# wide format
knitr::kable(data_long %>%  
               filter(id == 4) %>% 
               select(id, discipline, consensus, precision),
             caption = "Long format data")
```

### Date

```{r}
d <- d |> 
  mutate(date = format(as.Date(StartDate), "%B %d, %Y")) 
```

### Writing cleaned data file

```{r}
# store data
write_csv(d, "data/cleaned_wide.csv")
write_csv(data_long, "data/cleaned_long.csv")

# to prevent factor levels, save as RDS
saveRDS(d, "data/cleaned_wide.rds")
saveRDS(data_long, "data/cleaned_long.rds")

```
